\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-01-30 README.Rmd
cleaned up ERSCP CIF's in 2x2 grid. Need to show
equivalence between survreg and casebase in comparison of coefficients

Source code for Casebase paper

for template, use

```R 
devtools::install_github("rstudio/rticles")
```

>>>>

Source code for casebase paper. Install the following packages from GitHub

```R 
# install.packages("devtools")

devtools::install_github("rstudio/rticles") # for template
devtools::install_github("benmarwick/rrtools") # for building the research compendium/docker container
```

# To contribute

1. Modify the `analysis/paper/paper.Rmd` file.
2. Run `rrtools::add_dependencies_to_description()`. This command will ensure all used packages are added to the DESCRIPTION file.
3. Compile the `paper.Rmd` file
4. Commit and push your changes to GitHub

### 2020-01-30 analysis/paper/paper.Rmd
cleaned up ERSCP CIF's in 2x2 grid. Need to show
equivalence between survreg and casebase in comparison of coefficients

\begin{table}
  \begin{tabular}{p{1.8in}lll}
\hline
    &  Parameters &  Density \proglang{R} function & \code{dist}\\
    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
\hline
    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
\hline
  \end{tabular}
  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
  \label{tab:dists}
\end{table}

@@

```{r plot-erspc-data, eval = eval_cs1}
pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
plot(pt_object)
```

@@

```{r plot-stratified-erspc-data, eval = eval_cs1}
pt_object_strat <- casebase::popTime(ERSPC, 
                                     event = "DeadOfPrCa", 
                                     exposure = "ScrArm")
plot(pt_object_strat)
```

@@

```{r erspc-cox-cif, eval = eval_cs1}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
                       ignore = 99)

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence (%)", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]), 
     yscale = 100)
legend("topleft", 
       legend = c("Control group", "Screening group"), 
       col = c("red","blue"),
       lty = c(1, 1), 
       bg = "gray90")

casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
                                                  data = ERSPC, 
                                                  ratio = 100)
smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
                                          time = seq(0,15,0.1), 
                                          newdata = new_data)
lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)

casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
                                 data = ERSPC, 
                                 ratio = 100)
smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
                                          time = seq(0,15,0.1), 
                                          newdata = new_data)
lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)

casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                                    data = ERSPC, 
                                    ratio = 100)
lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)

legend("topleft", 
       legend = c("Control group (Cox)","Control group (Casebase)",
                  "Screening group (Cox)", "Screening group (Casebase)"), 
       col = c("red","red", "blue","blue"),
       lty = c(1, 2, 1, 2), 
       bg = "gray90")

```

@@

```{r erspc-casebase-cif, eval = eval_cs1}

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence (%)", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]),
     yscale = 100)

```

@@

```{r erspc-casebase-weibull, eval = eval_cs1}

summary(casebase_time)
exp(coef(casebase_time))
exp(confint(casebase_time))
```

@@

```{r erspc-casebase-weibull-cif, eval = eval_cs1}

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence (%)", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]),
     yscale = 100)

legend("topleft", 
       legend = c("Control group (Cox)","Control group (Casebase)",
                  "Screening group (Cox)", "Screening group (Casebase)"), 
       col = c("red","red", "blue","blue"),
       lty = c(1, 2, 1, 2), 
       bg = "gray90")
```

@@

```{r erspc-casebase-splines, eval = eval_cs1}

summary(casebase_splines)
exp(coef(casebase_splines))
exp(confint(casebase_splines))
```

@@

```{r erspc-casebase-splines-cif, eval = eval_cs1}
smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
                                    time = seq(0,15,0.1), 
                                    newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence (%)", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]),
     yscale = 100)

legend("topleft", 
       legend = c("Control group (Cox)","Control group (Casebase)",
                  "Screening group (Cox)", "Screening group (Casebase)"), 
       col = c("red","red", "blue","blue"),
       lty = c(1, 2, 1, 2), 
       bg = "gray90")
```

@@

```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group",])
CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group",])
CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group",])

table_ci <- rbind(CI_cox, CI_exp, CI_time, CI_splines)
colnames(table_ci) <- c("HR", "LB", "UB")

data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>% 
  bind_cols(as.data.frame(table_ci)) %>% 
  mutate(CI = paste0("(", round(LB, 2), ", ",
                     round(UB, 2), ")"),
         Width = UB - LB) %>% 
  dplyr::select(-LB, -UB) %>% 
  knitr::kable(format = "latex", booktabs = TRUE)  %>%
  kable_styling()
```

@@

```{r erscp-compare, eval = eval_cs1}
anova(casebase_exponential, casebase_splines, test = "LRT")
```

@@

```{r erscp-aic, eval = eval_cs1}
c("Exp." = AIC(casebase_exponential),
  "Weib." = AIC(casebase_time),
  "Splines" = AIC(casebase_splines))
```

>>>>

<!--
\begin{table}
  \begin{tabular}{p{1.8in}lll}
\hline
    &  Parameters &  Density \proglang{R} function & \code{dist}\\
    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
\hline
    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
\hline
  \end{tabular}
  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
  \label{tab:dists}
\end{table}
-->

@@

```{r plot-erspc-data, eval = eval_cs1}
# pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
# plot(pt_object)
```

@@

```{r plot-stratified-erspc-data, eval = eval_cs1}
pt_object_strat <- casebase::popTime(ERSPC, 
                                     event = "DeadOfPrCa", 
                                     exposure = "ScrArm")
# plot(pt_object_strat)
```

@@

```{r erspc-cox-cif, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap="Cumulative incidence functions for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression. (A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution. (B) Linear function of time which is equivalent to a Gompertz distribution. (C) The natural logarithm of time which is equivalent to a Weibull distribution. (D) Cubic b-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]
new_data <- data.frame(ScrArm = c("Control group","Screening group"), ignore = 99)

fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))

fits <- lapply(fmla, function(i) {
  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
})

risks <- lapply(fits, function(i) {
  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
})

# casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
#                                                   data = ERSPC, 
#                                                   ratio = 100)
# smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
#                                           time = seq(0,15,0.1), 
#                                           newdata = new_data)
# 
# casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
#                                            data = ERSPC, 
#                                            ratio = 100)
# smooth_risk_weibull <- casebase::absoluteRisk(object = casebase_weibull, 
#                                            time = seq(0,15,0.1), 
#                                            newdata = new_data)
# casebase_splines <- casebase::fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
#                                               data = ERSPC, 
#                                               ratio = 100)
# smooth_risk_splines <- casebase::absoluteRisk(object = casebase_splines, 
#                                               time = seq(0,15,0.1), 
#                                               newdata = new_data)
# 
# casebase_interaction <- casebase::fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time*ScrArm, 
#                                                   data = ERSPC, 
#                                                   ratio = 100)
# smooth_risk_interaction <- casebase::absoluteRisk(object = casebase_interaction, 
#                                                   time = seq(0,15,0.1), 
#                                                   newdata = new_data)

# par(mfrow = c(2, 2))
# casebase::hazardPlot(casebase_interaction, newdata = new_data[2,], ci = FALSE)
# casebase::hazardPlot(casebase_interaction, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)
# 
# casebase::hazardPlot(casebase_splines, newdata = new_data[2,], ci = FALSE)
# casebase::hazardPlot(casebase_splines, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)
# 
# casebase::hazardPlot(casebase_time, newdata = new_data[2,], ci = FALSE)
# casebase::hazardPlot(casebase_time, newdata = new_data[1,], add = TRUE, lty = 2)
# 
# casebase::hazardPlot(casebase_exponential, newdata = new_data[2,], ci = FALSE)
# casebase::hazardPlot(casebase_exponential, newdata = new_data[1,], add = TRUE, lty = 2)

# dev.off()
layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
par(mar = c(0, 4.1, 4.1, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",
     xlab = "",
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
     #                exp(coef(cox_model)), 
     #                exp(confint(cox_model))[1], 
     #                exp(confint(cox_model))[2]), 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Casebase)",
                  "Screening (Cox)", "Screening (Casebase)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = 15*0.97, y = 0.05/100, labels = "(A)", cex = 1.5)

par(mar = c(4.1, 4.1, 0, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "", xlab = "",
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
text(x = 15*0.97, y = 0.05/100, labels = "(C)", cex = 1.5)
mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)

par(mar = c(0, 0, 4.1, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
     xaxt = 'n', yaxt = 'n',
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = 15*0.97, y = 0.05/100, labels = "(B)", cex = 1.5)

par(mar = c(4.1, 0, 0, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
     yaxt = 'n',
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = 15*0.97, y = 0.05/100, labels = "(D)", cex = 1.5)
mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
```

## Equivalence between casebase and AFT

```{r}
# exponential -------------------------------------------------------------

casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
                                                  data = ERSPC, 
                                                  ratio = 100)
aft_exponential <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
                           data = ERSPC, dist = "exp")

-coef(aft_exponential)
coef(casebase_exponential)

# weibull -----------------------------------------------------------------

casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
                                           data = ERSPC, 
                                           ratio = 100)
aft_weibull <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
                       data = ERSPC, dist = "weibull")

-coef(aft_weibull)/aft_weibull$scale
coef(casebase_weibull)
```

```{r, eval = F}
knitr::knit_exit()
```

@@

@@

@@

@@

@@

@@

```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
CI_exp <- c(exp(coef(fits$exponential))["ScrArmScreening group"], exp(confint(fits$exponential))["ScrArmScreening group",])
CI_gompertz <- c(exp(coef(fits$gompertz))["ScrArmScreening group"], exp(confint(fits$gompertz))["ScrArmScreening group",])
CI_weibull <- c(exp(coef(fits$weibull))["ScrArmScreening group"], exp(confint(fits$weibull))["ScrArmScreening group",])
CI_splines <- c(exp(coef(fits$splines))["ScrArmScreening group"], exp(confint(fits$splines))["ScrArmScreening group",])

table_ci <- rbind(CI_cox, CI_exp, CI_gompertz, CI_weibull, CI_splines)
colnames(table_ci) <- c("HR", "LB", "UB")

data.frame(Model = c("Cox", "Exponential", "Gompertz", "Weibull", "Splines")) %>% 
  bind_cols(as.data.frame(table_ci)) %>% 
  mutate(CI = paste0("(", round(LB, 2), ", ",
                     round(UB, 2), ")"),
         Width = UB - LB) %>% 
  dplyr::select(-LB, -UB) %>% 
  knitr::kable(format = "latex", booktabs = TRUE)  %>%
  kable_styling()
```

@@

```{r erscp-compare, eval = eval_cs1}
anova(fits$exponential, fits$splines, test = "LRT")
```

@@

```{r erscp-aic, eval = eval_cs1}
c("Exp." = AIC(fits$exponential),
  "Gomp." = AIC(fits$gompertz),
  "Weib." = AIC(fits$exponential),
  "Splines" = AIC(fits$splines))
```
